---
title: "Data Science Project 2 -  Return to Education and Income in the United States (2023)"
author:
- Constantine Zhang
- Ha Nguyen
- Kemal Ege Sural
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---
```{r library, include = FALSE}
library(tidyverse)
library(haven)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(results="markup", warning = F, message = F,
                      fig.width=8, fig.height=6, fig.dim=c(8,6), fig.align='center')
options(scientific=T, digits = 3,scipen = 999)
```
```{r data, include = FALSE}
acs2023 <- read.csv("acs2023.csv")
```
# Literature Review - Ha

```{r }
acs_raw <- read.csv("acs2023.csv")
df <- acs_raw # copy


#Renaming

names(df)[names(df) == "inctot"]           <- "total_income"
names(df)[names(df) == "real_gdp"]         <- "county_gdp"
names(df)[names(df) == "real_pc_income"]   <- "county_pc_income"
names(df)[names(df) == "total_employment"] <- "county_total_jobs"

names(df)[names(df) == "educ"]     <- "education_code"
names(df)[names(df) == "age"]      <- "age"              
names(df)[names(df) == "sex"]      <- "gender_code"
names(df)[names(df) == "race"]     <- "race_code"
names(df)[names(df) == "marst"]    <- "marital_status_code"
names(df)[names(df) == "nchild"]   <- "num_children"
names(df)[names(df) == "citizen"]  <- "citizenship_code"      
names(df)[names(df) == "empstat"]  <- "employment_status_code" 
names(df)[names(df) == "labforce"] <- "labor_force_code"       
names(df)[names(df) == "perwt"]    <- "person_weight"

#Factor changes

df$gender <- factor(df$gender_code, 
                    levels = c(1, 2), 
                    labels = c("Male", "Female"))

df$marital_status <- factor(df$marital_status_code, levels = 1:6, 
                            labels = c("Married (Present)", "Married (Absent)", 
                                       "Separated", "Divorced", "Widowed", "Single"))


df$race <- factor(df$race_code, levels = 1:6,
                  labels = c("White", "Black", "Native American", 
                             "Chinese", "Japanese", "Other Asian/Pacific"))

df$citizenship <- factor(df$citizenship_code, levels = 1:3,
                         labels = c("Born Abroad (US Parents)", 
                                    "Naturalized Citizen", 
                                    "Not a Citizen"))

df$employment_status <- factor(df$employment_status_code, levels = 1:3,
                               labels = c("Employed", "Unemployed", "Not in Labor Force"))

df$labor_force <- factor(df$labor_force_code, levels = 1:2,
                         labels = c("Not in Labor Force", "In Labor Force"))


df$education_level <- factor(df$education_code, levels = 1:11,
                             labels = c("Nursery-Gr4", "Gr5-8", "Gr9", "Gr10", "Gr11", "Gr12",
                                        "1yr College", "2yr College", "3yr College", "4yr College", "5+yr College"))


df$educ_group <- NA
df$educ_group[df$education_code <= 6] <- "HS or Less"
df$educ_group[df$education_code %in% c(7, 8, 9)] <- "Some College"
df$educ_group[df$education_code == 10] <- "Bachelor's"
df$educ_group[df$education_code == 11] <- "Post-Grad"
df$educ_group <- factor(df$educ_group, levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"))

print(names(df))

#Removing NA/ 0 INCOME
df <- na.omit(df)
df <- df[df$total_income > 0, ]
```




# EDA - based on the smart questions in the proposal - for this survey, we will can calculate state level data + individual level data, try to think in both directions - deadline - before thanksgiving (Nov 27)


1. Statistics summary, plot 

```{r }

#  Income distribution
par(mfrow = c(1, 1)) 

hist(df$total_income, 
     breaks = 100, 
     xlim = c(0, 350000),
     main = "Distribution of Annual Income",
     xlab = "Annual Income ($)",
     ylab = "Frequency",
     col = "cadetblue",
     border = "white")

# Income by Education level
boxplot(total_income ~ educ_group, data = df,
        main = "Income Ranges by Education Level",
        xlab = "Education Level",
        ylab = "Annual Income ($)",
        col = "lightblue",
        outline = FALSE, 
        las = 1) 

# Median Income
income_by_educ <- aggregate(total_income ~ educ_group, data = df, FUN = median)
print(income_by_educ)

# Gender income difference by education
par(mar = c(8, 4, 4, 2))
boxplot(total_income ~ gender * educ_group, data = df,
        main = "Gender Pay Gap by Education",
        xlab='',
        ylab = "Annual Income ($)",
        col = c("cornflowerblue", "pink"), # Mavi=Erkek, Pembe=Kadın
        outline = FALSE,
        las = 2, 
        cex.axis = 0.7) # Yazı boyutunu biraz küçült

legend("topleft", legend = c("Male", "Female"), 
       fill = c("cornflowerblue", "pink"), bty = "n")

# Categoric
par(mfrow = c(1, 2)) 

# Education
educ_counts <- table(df$educ_group)
barplot(educ_counts, 
        main = "Sample Size by Education",
        col = "orange",
        las = 2,
        cex.names = 0.7)

# Race
race_counts <- sort(table(df$race), decreasing = FALSE) 
par(mar=c(5, 6, 4, 1)) 
barplot(race_counts, 
        main = "Racial Composition", 
        xlim = c(0,100000),
        horiz = TRUE, 
        col = "lightgray",
        las = 1,
        cex.names = 0.7)
```

2. Normal distribution test

3. Hypothesis testing 


# ANOVA (Analysis of Variance) - FOR EDUCATION LEVELS
 
 Research Question: Does education level result in a statistically significant difference in income?
 
 H0 (Null Hypothesis): The mean income of all education groups is equal.
 
 H1 (Alternative Hypothesis): At least one group's income is different from the others.

```{r }

anova <- aov(log(total_income) ~ educ_group, data = df)

print(summary(anova))
```

A one-way ANOVA test was conducted to determine if education level creates a statistically significant difference in income. The results indicate a strong relationship, with a high F-value ($13,868$) and a p-value of less than 0.001. Because the p-value is extremely low, we reject the null hypothesis and confirm that income levels vary significantly across education groups. However, since the ANOVA does not specify which groups differ, a post-hoc test is required to identify the specific differences.

```{r }
tukey<- TukeyHSD(anova)

print(tukey)

par(mar=c(5, 15, 4, 2)) 
plot(tukey, las=1, col="red")

```

The results show that every single pairwise comparison is statistically significant. This confirms a clear hierarchy in earnings: those with 'High School or Less' earn the lowest, followed by 'Some College' and 'Bachelor's', while the 'Post-Grad' group has the highest income. In short, each step up in education level results in a statistically significant increase in total income.

# T-Test (Independent Samples) - FOR GENDER
 
 Research Question: Is there a difference between the mean incomes of males and females?
 
 H0 (Null Hypothesis): The mean incomes of males and females are EQUAL.
 
 H1 (Alternative Hypothesis): The mean incomes of males and females are NOT EQUAL.


```{r }

t_test <- t.test(log(total_income) ~ gender, data = df)
print(t_test)
```

A Welch Two Sample t-test was conducted to compare the average incomes of male and female groups. The results indicate a highly significant difference between the two genders ($t = 88, p < 0.001$). The analysis shows that the average log-income for males ($10.8$) is higher than the average log-income for females ($10.3$). Additionally, the 95% confidence interval is entirely positive ($0.46$ to $0.48$), which confirms that males consistently earn more than females in this dataset. Based on this, the null hypothesis is rejected.


# Pearson Correlation Test - income vs county gdp

Research Question: To what extent is an individual's total income associated with the economic prosperity (GDP) of their county?

H0 (Null Hypothesis): There is NO linear correlation between individual income and county-level GDP.
 
H1 (Alternative Hypothesis): There is a statistically significant linear correlation between individual income and county-level GDP.

```{r }
cor_test <- cor.test(log(df$total_income), log(df$county_gdp), method = "pearson")
print(cor_test)

```

A Pearson correlation test was conducted to examine the relationship between county-level GDP and individual total income. 

The results show a statistically significant relationship ($t = 14, p < 0.001$). However, the correlation coefficient is extremely low ($r = 0.03$). This indicates that although there is a positive association between the wealth of a region and personal income, the relationship is negligible. In other words, living in a richer county has almost no direct impact on an individual's income in this dataset.

# Regression

1. Plot of correlation

2. Run regression

3. Test the fit of the regression

- Confusion matrix

- Mcfaddle

- Anova Test between models (remove controlling variables)

- ROC/AUC

# Machine learning