---
title: "Data Science Project 2 -  Return to Education and Income in the United States (2023)"
author:
- Constantine Zhang
- Ha Nguyen
- Kemal Ege Sural
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r library, include = FALSE}
library(tidyverse)
library(haven)
library(corrplot)
library(ggplot2)
library(reshape2)
library(lsr)
library(rpart)
library(interactions)
library(patchwork)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(results="markup", warning = F, message = F,
                      fig.width=8, fig.height=6, fig.dim=c(8,6), fig.align='center')
options(scientific=T, digits = 3,scipen = 999)
```

```{r data, include = FALSE}
acs2023 <- read.csv("acs2023.csv")
```
# Literature Review - Ha

```{r }
acs_raw <- read.csv("acs2023.csv")
df <- acs_raw # copy


#Renaming

names(df)[names(df) == "inctot"]           <- "total_income"
names(df)[names(df) == "real_gdp"]         <- "state_real_gdp"
names(df)[names(df) == "real_pc_income"]   <- "state_pc_income"
names(df)[names(df) == "total_employment"] <- "state_total_jobs"

names(df)[names(df) == "educ"]     <- "education"
names(df)[names(df) == "age"]      <- "age"              
names(df)[names(df) == "sex"]      <- "gender"
names(df)[names(df) == "race"]     <- "race"
names(df)[names(df) == "marst"]    <- "marital_status"
names(df)[names(df) == "nchild"]   <- "num_children"
names(df)[names(df) == "citizen"]  <- "citizenship"      
names(df)[names(df) == "empstat"]  <- "employment_status" 
names(df)[names(df) == "labforce"] <- "labor_force"       
names(df)[names(df) == "perwt"]    <- "person_weight"

#Factor changes

df$gender <- factor(df$gender, 
                    levels = c(1, 2), 
                    labels = c("Male", "Female"))

df$marital_status <- factor(df$marital_status, levels = 1:6, 
                            labels = c("Married (Present)", "Married (Absent)", 
                                       "Separated", "Divorced", "Widowed", "Single"))


df$race <- factor(df$race, levels = 1:6,
                  labels = c("White", "Black", "Native American", 
                             "Chinese", "Japanese", "Other Asian/Pacific"))

df$citizenship <- factor(df$citizenship, levels = 1:3,
                         labels = c("Born Abroad (US Parents)", 
                                    "Naturalized Citizen", 
                                    "Not a Citizen"))

df$employment_status <- factor(df$employment_status, levels = 1:3,
                               labels = c("Employed", "Unemployed", "Not in Labor Force"))

df$labor_force <- factor(df$labor_force, levels = 1:2,
                         labels = c("Not in Labor Force", "In Labor Force"))


df$education_level <- factor(df$education, levels = 1:11,
                             labels = c("Nursery-Gr4", "Gr5-8", "Gr9", "Gr10", "Gr11", "Gr12",
                                        "1yr College", "2yr College", "3yr College", "4yr College", "5+yr College"))


df$educ_group <- NA
df$educ_group[df$education <= 6] <- "HS or Less"
df$educ_group[df$education %in% c(7, 8, 9)] <- "Some College"
df$educ_group[df$education == 10] <- "Bachelor's"
df$educ_group[df$education == 11] <- "Post-Grad"
df$educ_group <- factor(df$educ_group, levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"))

print(names(df))

#Removing NA/ 0 INCOME
df <- na.omit(df)
df <- df[df$total_income > 0, ]
```




# DEA - based on the smart questions in the proposal - for this survey, we can calculate state level data + individual level data, try to think in both directions - deadline - before thanksgiving (Nov 27)


1. Statistics summary, plot 

```{r }

#  Income distribution
par(mfrow = c(1, 1)) 

hist(df$total_income, 
     breaks = 100, 
     xlim = c(0, 350000),
     main = "Distribution of Annual Income",
     xlab = "Annual Income ($)",
     ylab = "Frequency",
     col = "cadetblue",
     border = "white")
```

The histogram is right-skewed, showing that most incomes are low while a few high-income outliers form a long tail to the right. Because these outliers pull the mean upward, the median is a more accurate measure of the typical income.

```{r }
# Income by Education level
boxplot(total_income ~ educ_group, data = df,
        main = "Income Ranges by Education Level",
        xlab = "Education Level",
        ylab = "Annual Income ($)",
        col = "lightblue",
        outline = FALSE, 
        las = 1) 

# Median Income
income_by_educ <- aggregate(total_income ~ educ_group, data = df, FUN = median)
print(income_by_educ)
```

The boxplots shows a clear positive link between education and income. As education levels rise, the median income consistently increases. However, the variance of salaries also gets much wider with higher education. While high school graduates earn within a narrow, consistent range, the "Post-Grad" group has a very tall box and long whiskers, indicating a huge gap between the lowest and highest earners in that category.

```{r }
# Gender income difference by education
par(mar = c(8, 4, 4, 2))
boxplot(total_income ~ gender * educ_group, data = df,
        main = "Gender Pay Gap by Education",
        xlab='',
        ylab = "Annual Income ($)",
        col = c("cornflowerblue", "pink"), # Mavi=Erkek, Pembe=Kadın
        outline = FALSE,
        las = 2, 
        cex.axis = 0.7) # Yazı boyutunu biraz küçült

legend("topleft", legend = c("Male", "Female"), 
       fill = c("cornflowerblue", "pink"), bty = "n")


```

The graph compares incomes for men and women across different education levels. It clearly shows that in every category, men have a higher median income than women. This pay gap gets much wider as education levels rise. While the difference is smaller for high school graduates, it becomes very large for the "Post-Grad" group, where men earn significantly more than women even with the same level of education.

```{r }
# Categoric
par(mfrow = c(1, 2)) 

# Education
educ_counts <- table(df$educ_group)
barplot(educ_counts, 
        main = "Sample Size by Education",
        col = "orange",
        las = 2,
        cex.names = 0.7)


```

This chart counts how many people in the study have different levels of education. The biggest group is people with a high school education or less. The smallest group is people with only some college experience.

```{r }
# Race
race_counts <- sort(table(df$race), decreasing = FALSE) 
par(mar=c(5, 6, 4, 1)) 
barplot(race_counts, 
        main = "Racial Composition", 
        xlim = c(0,100000),
        horiz = TRUE, 
        col = "lightgray",
        las = 1,
        cex.names = 0.7)
```

This chart shows the race of the people in the study. The vast majority of people are White or from the "Other Asian/Pacific" group. On the other hand, there are very few Native American or Japanese participants.


2. Normal distribution test

3. Hypothesis testing 


# ANOVA (Analysis of Variance) - FOR EDUCATION LEVELS
 
 Research Question: Does education level result in a statistically significant difference in income?
 
 H0 (Null Hypothesis): The mean income of all education groups is equal.
 
 H1 (Alternative Hypothesis): At least one group's income is different from the others.

```{r }

anova <- aov(log(total_income) ~ educ_group, data = df)

print(summary(anova))
```

A one-way ANOVA test was conducted to determine if education level creates a statistically significant difference in income. The results indicate a strong relationship, with a high F-value ($13,868$) and a p-value of less than 0.001. Because the p-value is extremely low, we reject the null hypothesis and confirm that income levels vary significantly across education groups. However, since the ANOVA does not specify which groups differ, a post-hoc test is required to identify the specific differences.

```{r }
tukey<- TukeyHSD(anova)

print(tukey)

par(mar=c(5, 15, 4, 2)) 
plot(tukey, las=1, col="red")

```

The results show that every single pairwise comparison is statistically significant. This confirms a clear hierarchy in earnings: those with 'High School or Less' earn the lowest, followed by 'Some College' and 'Bachelor's', while the 'Post-Grad' group has the highest income. In short, each step up in education level results in a statistically significant increase in total income.

# T-Test (Independent Samples) - FOR GENDER
 
 Research Question: Is there a difference between the mean incomes of males and females?
 
 H0 (Null Hypothesis): The mean incomes of males and females are EQUAL.
 
 H1 (Alternative Hypothesis): The mean incomes of males and females are NOT EQUAL.


```{r }

t_test <- t.test(log(total_income) ~ gender, data = df)
print(t_test)
```

A Welch Two Sample t-test was conducted to compare the average incomes of male and female groups. The results indicate a highly significant difference between the two genders ($t = 88, p < 0.001$). The analysis shows that the average log-income for males ($10.8$) is higher than the average log-income for females ($10.3$). Additionally, the 95% confidence interval is entirely positive ($0.46$ to $0.48$), which confirms that males consistently earn more than females in this dataset. Based on this, the null hypothesis is rejected.


# Pearson Correlation Test - income vs state gdp

Research Question: To what extent is an individual's total income associated with the economic prosperity (GDP) of their state?

H0 (Null Hypothesis): There is NO linear correlation between individual income and state-level GDP.
 
H1 (Alternative Hypothesis): There is a statistically significant linear correlation between individual income and state-level GDP.

```{r }
cor_test <- cor.test(log(df$total_income), log(df$state_real_gdp), method = "pearson")
print(cor_test)

```

A Pearson correlation test was conducted to examine the relationship between state-level GDP and individual total income. 

The results show a statistically significant relationship ($t = 14, p < 0.001$). However, the correlation coefficient is extremely low ($r = 0.03$). This indicates that although there is a positive association between the wealth of a region and personal income, the relationship is negligible. In other words, living in a richer state has almost no direct impact on an individual's income in this dataset.

# Regression

1. Plot of correlation

To explore main explanatory variable education level against main dependent variable real per capital income, we fiest look at the correlation plot between these two variables.  

Because our main explanatory variable is categorical variable, we be using correlation matrix for this section. 

For numerical variable and categorical, we need to use Eta method to determine the correlation.

```{r}
etaSquared(aov( total_income ~ education_level, data = df))
eta_value <- function(cat, num) {
  as.numeric(etaSquared(aov(num ~ cat))[1])
}

eta_value(df$education_level, df$total_income)
```

This result imply that education level explains 14.1% of the variation in total income. This is on the threshold of large effect which means education level has a meaningful and strong association with persoanl total income. 

2. Run regression

```{r change level to factor and reference level}
df$education_level <- factor(df$educ_group)
df$race <- factor(df$race)
df$gender <- factor(df$gender)
df$race <- relevel(df$race, ref = "White")
df$gender <- relevel(df$gender, ref = "Male")
df$educ_group <- relevel(df$educ_group, ref = "HS or Less")
```



```{r linear regression without control}
model.base <- lm( log(total_income) ~ educ_group, data = df)
summary(model.base)
```
Our regression estimation for simple linear regression is:

total income 2023(personal level) = 26928 + 1584*Grade5-8 + 3312*Grade9 + 3276*Grade10 + 1269*Grade11 + 12964*Grade12 + 20261*1 year College + 25400*2 year college + 57307*4 year college + 106864* 5 or more year college 


Based on the simple regression model, education have a higher and higher influence on personal level income after 12 grade and all of those coefficients are significant. This results correpsondes to the real world because law forbide child labor thus income would only be associate with education level at least after high school when a person could work legaly full time. 

```{r linear regression with control for gender, age, and race}
model.gender.race <- lm( log(total_income) ~ educ_group + gender + race + age , data = df)
summary(model.gender.race)
```
This specification with age, gender, and race as control has a higher R squared value. We also get significance from all the coefficient. However, this specification only allow change in intercept not slope between different race and gender. 

```{r vif check model.gender.race}
car::vif(model.gender.race)
```

```{r linear regression with interaction terms for gender, age, and race}
model.interaction <- lm( log(total_income) ~  educ_group   * gender + educ_group* race + gender * race + age  , data = df)
summary(model.interaction)
car::vif(model.interaction)
```
Our third regression model specification allow change of slope between race and gender. The R squared of this model is slightly higher, but we still get significance in some of the coefficients due to the substantially large number of observations we have in the data frame. 


```{r}
df$educ_group <- factor(df$educ_group,
                        levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"),
                        ordered = TRUE)
levels(df$educ_group)


```
To better understanding and interpreting the result coefficients, we employ interaction plots for the third specification. 

```{r interaction plot gender}
newdata <- expand.grid(
  educ_group = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"),
  gender = unique(df$gender),
  race = unique(df$race),
  age = mean(df$age, na.rm = TRUE)  
)


newdata$educ_group <- factor(newdata$educ_group,levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"), ordered = TRUE)
predictions <- predict(model.interaction, newdata, se.fit = TRUE, interval = "confidence")

pred_df %>%
  group_by(educ_group, gender) %>%
  summarise(
    log_income_pred = mean(log_income_pred),
    ci_lower = mean(ci_lower),
    ci_upper = mean(ci_upper)
  ) -> pred_summary



ggplot(pred_summary, aes(x = educ_group, y = log_income_pred, color = gender)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
                width = 0.2,
                position = position_dodge(width = 0.5)) +
  scale_x_discrete(limits = c("HS or Less", "Some College", "Bachelor's", "Post-Grad")) +
  labs(x = "Education Level",
       y = "Predicted log(Income)",
       color = "Gender",
       title = "Predicted log(Income) by Education and Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This interaction plot represent the predicted value of log (income) for male and female based on their education level include confidence interval of 95%. This graph shows expectd results: as education level increase, income increase. Moreoevr, male have higher income compared to female in all education levels.  

```{r interaction plot race }
library(emmeans)
race_emm <- emmeans(model.interaction, ~ race)  
race_df <- as.data.frame(race_emm)
colnames(race_df)
ggplot(race_df, aes(x = race, y = emmean)) +
  geom_point(size = 3, color = "darkgreen") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, color = "darkgreen") +
  labs(x = "Race",
       y = "Predicted log(Income)",
       title = "Predicted log(Income) by Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

This interaction plot demonstrate the predicted value for log(income) based on race, the reusult is also expected as white has the highest predicted income while native american and african american has the lowest predicted income.

```{r race and gender plot}
race_gender_emm <- emmeans(model.interaction, ~ race * gender)
race_gender_df <- as.data.frame(race_gender_emm)

race_gender_df$race <- factor(race_gender_df$race, levels = levels(df$race))

ggplot(race_gender_df, aes(x = race, y = emmean, color = gender)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.2,
                position = position_dodge(width = 0.5)) +
  labs(x = "Race",
       y = "Predicted log(Income)",
       color = "Gender",
       title = "Predicted log(Income) by Race and Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Next, we include graph for predicted log(income) based on both race and gender to see the gender wage gap between different races. Japanese has the largest gender income gap follow by white while black and Chinese has the smallest gender income gap. 

```{r race, gender, and education level}
pred_emm <- emmeans(model.interaction, ~ educ_group * gender * race)
pred_df <- as.data.frame(pred_emm)


pred_df$educ_group <- factor(pred_df$educ_group,
                             levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"),
                             ordered = TRUE)
pred_df$race <- factor(pred_df$race, levels = levels(df$race))
pred_df$gender <- factor(pred_df$gender, levels = levels(df$gender))

ggplot(pred_df, aes(x = educ_group, y = emmean, color = gender)) +
  geom_point(position = position_dodge(width = 0.5), size = 1.5) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.2,
                position = position_dodge(width = 0.5)) +
  facet_wrap(~ race) +
  labs(x = "Education Level",
       y = "Predicted log(Income)",
       color = "Gender",
       title = "Predicted log(Income) by Education, Gender, and Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
This interaction plot combines everything together plotting predicted log(income) based on gender, race, and education level. It is clear that log(income) and education have a positve correlation in every racial group and gender income gap also presist in every racial group. 


3. Test the fit of the regression

- Anova Test between models (remove controlling variables)

```{r test for base model and control variable model}
anova(model.base, model.gender.race)
```
ANOVA test between base model and model with control variables shows a p value close to 0 meaning control variables of age, gender, and race significantly improve the model.


```{r test control model and interaction model}
anova(model.gender.race, model.interaction)
```
ANOVA test between model with control variables and model inetraction terms shows a p value close to 0 meaning adding interaction terms and allowing the slope to changed between races and gender significantly improve the model.


# Tree-based model

## Grow the tree by each level of education attain

Convert education to categorical so rpart can split them properly.

- Outcome: total_income (continuous)

- Model type: regression tree (uses MSE minimization) as between continous dependent variable and mixed independent variable

- Predictors:

  - Demographics: education, gender, age, marital status, race, citizenship

  - Family: number of children

  - State macro controls: log GDP, log jobs

  - maxdepth = 7: tree can grow up to 7 levels deep
  
```{r tree 1}
set.seed(1)

df <- df %>%
  mutate(education_cat = factor(education))

summary(df$education_cat)
```
```{r tree 2}
income_regression <- rpart(total_income ~ education_cat + num_children + gender + age + marital_status + race + citizenship + log(state_real_gdp) + log(state_total_jobs), data=df, method="anova", control = list(maxdepth = 7) )
```

```{r summary tree}
summary(income_regression) 
```
1. CP Table (How Much Each Split Helps)

CP nsplit rel error xerror    xstd
1 0.1085      0     1.000  1.000 0.01010
2 0.0302      1     0.892  0.892 0.00927
3 0.0172      2     0.861  0.861 0.00909
4 0.0100      3     0.844  0.844 0.00891


Start error = 1.000

After first split: error → 0.892 (−11%)

After second split: error → 0.861

After third split: error → 0.844

Interpretation:

The first split is by far the strongest (education).

Additional splits reduce error but with much smaller gains.

2. Variable Importance

Education explains 88% of all the splitting power of the tree.

Gender adds 11%.

Everything else combined = 1%.

3. Root Node (Before Any Split)

**Node 1:**
n = 229,342
mean = 75,800
MSE = 1.02e+10

The average total income in the whole dataset is $75.8k.

MSE is large because incomes vary a lot (normal).

**Node 2:**

Observation: 172,509 obs (lower-education group)

Income levels: $56,700

Even within the "lower education" categories, education still matters.

Each level increments income:

HS < Some College < Bachelor’s < (but Bachelor’s still partly in Node 3 9,11 - 3 years college and 5+ years)

**Node 3:**

- Right son (Node 3): 56,833 obs (higher-education group)

- Node 3 mean = $134,000

- continue split with gender, improve = 0.0361

Interpretation:

Among highly educated people:

Gender becomes the next major income divider.

This means men and women with the same high education credentials earn significantly different incomes.

**Node 4:**

Lowest-education group → $41k average income
→ Probably HS or less + young or single

**Node 5:**

Mid-education → $84k
→ Maybe Bachelor’s or some college groups

**Node 6:**

High education + Female → $105k

**Node 7:**

High education + Male → $158k
→ This is the high-education/male cluster: highest earners.

```{r plot tree 1}
library(rpart.plot)
rpart.plot(
  income_regression,
  type = 2,          
  extra = 101,       
  under = TRUE,      
  faclen = 0,        
  digits = -3,
  tweak = 1.1        
)
```
**Overall Takeaway**

1. Education is the overwhelmingly strongest predictor of income.

It defines the first split and most of the structure.

2. Gender is the second-most important variable.

Especially for high education groups.

3. State-level economics barely affect individual income.

The tree almost never splits on log(state_real_gdp) or state_total_jobs.

## Prune the Tree

We can "prune" the tree to reduce variance. A plot can be made after pruning, showing the effects in terms of deviance or other statistics.

```{r prune}
library(rattle)

print(income_regression$cptable)
plotcp(income_regression)
income_prune<- prune(income_regression, cp = income_regression$cptable[3,"CP"])
rpart.plot(
  income_prune,
  type = 2,          
  extra = 101,       
  under = TRUE,      
  faclen = 0,        
  digits = -3,
  tweak = 1.1        
)
```
The graph shows the result with different number of terminal/leaf nodes on the x (tree size), and the cross-validated prediction error value of the model on the y. 

# Train/Test Split with Tree Pruning Evaluation”

Set 5 fold

1. First fold Cross-Validation (1-fold CV) for Pruning

```{r, results="markup"}
set.seed(1)
df$fold <- floor(runif(nrow(df), 1, 5))
test.set <- df[df$fold == 1, ]
train.set <- df[df$fold != 1, ]
```

```{r train}
treefit_fold <- rpart(total_income ~ education_cat + num_children + gender + age +
                     marital_status + race + citizenship +
                     log(state_real_gdp) + log(state_total_jobs),
                     data = train.set,
                     method = "anova",
                     control = list(maxdepth = 7))
rpart.plot(
  treefit_fold,
  type = 2,          
  extra = 101,       
  under = TRUE,      
  faclen = 0,        
  digits = -3,
  tweak = 1.1        
)
```

```{r predict}
pred <- predict(treefit_fold, test.set)
plot(pred,test.set$total_income, main="Predicted vs. Observed Total Income", xlab="Tree Predicted Salary", ylab="Observed Salary")
treefit_fold_error <- sum((pred - test.set$total_income)^2)
treefit_fold_error
```

```{r prune the tree}
treefit_fold_prune <- prune(treefit_fold,
                            cp = treefit_fold$cptable[3, "CP"])

pred2 <- predict(treefit_fold_prune, test.set)
plot(pred2,test.set$total_income, main="Predicted vs. Observed Total Income", xlab="Tree Predicted Salary", ylab="Observed Salary")
treefit_fold_prune_error <- sum((pred2 - test.set$total_income)^2)
treefit_fold_prune_error
```

Prune model has more error, but this is just the first fold!.

2. Testing all 5 CV

we split the dataset into 5 random groups. Pick one group as test set in turns, and the remaining four as training data, we can build five different models and score the models using the test data five times. 

```{r four fold cv}
library(rpart.plot)

set.seed(1)

K <- 5
df$fold_2 <- sample(1:K, size = nrow(df), replace = TRUE)

cv_results <- data.frame(
  fold         = 1:K,
  full_error   = NA_real_,
  pruned_error = NA_real_,
  cp_row3      = NA_real_,
  error_diff   = NA_real_,   
  compare      = NA_character_   
)

for (k in 1:K) {
  test.set  <- df[df$fold_2 == k, ]
  train.set <- df[df$fold_2 != k, ]
  
  treefit_k <- rpart(
    total_income ~ education_cat + num_children + gender + age +
      marital_status + race + citizenship +
      log(state_real_gdp) + log(state_total_jobs),
    data   = train.set,
    method = "anova",
    control = rpart.control(maxdepth = 7)
  )
  
  pred_full <- predict(treefit_k, newdata = test.set)
  sse_full  <- sum((pred_full - test.set$total_income)^2)
  cv_results$full_error[k] <- sse_full
  
  cp_row3 <- treefit_k$cptable[3,"CP"]
  cv_results$cp_row3[k] <- cp_row3
  
  treefit_k_pruned <- prune(treefit_k, cp = cp_row3)
  
  pred_pruned <- predict(treefit_k_pruned, newdata = test.set)
  sse_pruned  <- sum((pred_pruned - test.set$total_income)^2)
  cv_results$pruned_error[k] <- sse_pruned

  cv_results$error_diff[k] <- sse_pruned - sse_full
  
  if (sse_pruned < sse_full) {
    cv_results$compare[k] <- "Pruned < Full"   
  } else if (sse_pruned > sse_full) {
    cv_results$compare[k] <- "Pruned > Full"   
  } else {
    cv_results$compare[k] <- "Pruned = Full"  
  }
}


cv_results
```
Across all five folds of the cross-validation exercise, the pruned tree consistently produced a higher sum of squared errors than the full tree, indicating that pruning reduced predictive accuracy in this dataset. This outcome is expected because the fully grown tree is already extremely small, with only three splits.

We selected the subtree associated with the smallest xstd in the complexity table, leading to choosing the third row. Although this row provides the most stable cross-validated error estimate, it also corresponds to a simpler tree that eliminates the final split present in the full model. By removing an informative split, the pruned tree loses explanatory power and underfits relative to the full model. As a result, the pruned trees capture less variation in total income, leading to systematically worse out-of-sample performance. In this setting, pruning does not improve generalization; the full tree provides the best balance of fit and complexity for predictive accuracy.

Summary table across folds:

```{r error_summary}
error_summary <- data.frame(
  model    = c("Full tree", "Pruned tree"),
  mean_SSE = c(mean(cv_results$full_error),
               mean(cv_results$pruned_error))
)

error_summary
```

The summary table of average SSE across folds reinforces the cross-validation results: the full tree achieves a lower mean SSE compared to the pruned tree, indicating that the pruned model performs worse on average. This outcome aligns with the individual-fold results, where pruning consistently increased prediction error. Because the full tree in this dataset is already extremely small, further pruning-based on selecting the subtree with the smallest xstd-removes an informative split and forces the model into an overly simplified structure. As a result, the pruned tree systematically underfits and fails to explain as much variation in total income as the full tree. Overall, the mean SSE comparison confirms that pruning does not improve predictive accuracy in this case, and the full tree provides better out-of-sample performance.

# Reference

Autor, D. H. (2014). Skills, education, and the rise of earnings inequality among the other 99 percent. Science, 344(6186), 843–851. https://doi.org/10.1126/science.1251868

Card, D. (1999). The causal effect of education on earnings. In O. Ashenfelter & D. Card (Eds.), Handbook of labor economics (Vol. 3, pp. 1801–1863). Elsevier.

Moretti, E. (2004). Human capital externalities in cities. In J. V. Henderson & J.-F. Thisse (Eds.), Handbook of regional and urban economics (Vol. 4, pp. 2243–2291). Elsevier. https://doi.org/10.1016/S1574-0080(04)80014-4

Ruggles, S., Flood, S., Sobek, M., Backman, D., Cooper, G., Rivera Drew, J. A., Richards, S., Rodgers, R., Schroeder, J., & Williams, K. C. W. (2025). IPUMS USA: Version 16.0 [ACS 1-year microdata for 2023]. IPUMS. https://doi.org/10.18128/D010.V16.0

U.S. Bureau of Economic Analysis. (2023). State Annual Summary Statistics: Personal Income, GDP, Consumer Spending, Price Indexes, and Employment. U.S. Department of Commerce. Retrieved from https://apps.bea.gov/regional/downloadzip.htm

