---
title: "Data Science Project 2 -  Return to Education and Income in the United States (2023)"
author:
- Constantine Zhang
- Ha Nguyen
- Kemal Ege Sural
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---
```{r lirary, include = FALSE}
library(tidyverse)
library(haven)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(results="markup", warning = F, message = F,
                      fig.width=8, fig.height=6, fig.dim=c(8,6), fig.align='center')
options(scientific=T, digits = 3,scipen = 999)
```
```{r data, include = FALSE}
acs2023 <- read.csv("acs2023.csv")
```
# Literature Review - Ha

# DEA - based on the smart questions in the proposal - for this survey, we will can calculate state level data + individual level data, try to think in both directions - deadline - before thanksgiving (Nov 27)

1. Statistics summary, plot 

2. Normal distribution test

3. Hypothesis testing 

# Regression

1. Plot of correlation

2. Run regression

3. Test the fit of the regression

- Confusion matrix

- Mcfaddle

- Anova Test between models (remove controlling variables)

- ROC/AUC