---
title: "Data Science Project 2 -  Return to Education and Income in the United States (2023)"
author:
- Constantine Zhang
- Ha Nguyen
- Kemal Ege Sural
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---
```{r library, include = FALSE}
library(tidyverse)
library(haven)
library(corrplot)
library(ggplot2)
library(reshape2)
library(lsr)
library(rpart)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(results="markup", warning = F, message = F,
                      fig.width=8, fig.height=6, fig.dim=c(8,6), fig.align='center')
options(scientific=T, digits = 3,scipen = 999)
```

```{r data, include = FALSE}
acs2023 <- read.csv("acs2023.csv")
```
# Literature Review - Ha

```{r }
acs_raw <- read.csv("acs2023.csv")
df <- acs_raw # copy


#Renaming

names(df)[names(df) == "inctot"]           <- "total_income"
names(df)[names(df) == "real_gdp"]         <- "state_real_gdp"
names(df)[names(df) == "real_pc_income"]   <- "state_pc_income"
names(df)[names(df) == "total_employment"] <- "state_total_jobs"

names(df)[names(df) == "educ"]     <- "education"
names(df)[names(df) == "age"]      <- "age"              
names(df)[names(df) == "sex"]      <- "gender"
names(df)[names(df) == "race"]     <- "race"
names(df)[names(df) == "marst"]    <- "marital_status"
names(df)[names(df) == "nchild"]   <- "num_children"
names(df)[names(df) == "citizen"]  <- "citizenship"      
names(df)[names(df) == "empstat"]  <- "employment_status" 
names(df)[names(df) == "labforce"] <- "labor_force"       
names(df)[names(df) == "perwt"]    <- "person_weight"

#Factor changes

df$gender <- factor(df$gender, 
                    levels = c(1, 2), 
                    labels = c("Male", "Female"))

df$marital_status <- factor(df$marital_status, levels = 1:6, 
                            labels = c("Married (Present)", "Married (Absent)", 
                                       "Separated", "Divorced", "Widowed", "Single"))


df$race <- factor(df$race, levels = 1:6,
                  labels = c("White", "Black", "Native American", 
                             "Chinese", "Japanese", "Other Asian/Pacific"))

df$citizenship <- factor(df$citizenship, levels = 1:3,
                         labels = c("Born Abroad (US Parents)", 
                                    "Naturalized Citizen", 
                                    "Not a Citizen"))

df$employment_status <- factor(df$employment_status, levels = 1:3,
                               labels = c("Employed", "Unemployed", "Not in Labor Force"))

df$labor_force <- factor(df$labor_force, levels = 1:2,
                         labels = c("Not in Labor Force", "In Labor Force"))


df$education_level <- factor(df$education, levels = 1:11,
                             labels = c("Nursery-Gr4", "Gr5-8", "Gr9", "Gr10", "Gr11", "Gr12",
                                        "1yr College", "2yr College", "3yr College", "4yr College", "5+yr College"))


df$educ_group <- NA
df$educ_group[df$education <= 6] <- "HS or Less"
df$educ_group[df$education %in% c(7, 8, 9)] <- "Some College"
df$educ_group[df$education == 10] <- "Bachelor's"
df$educ_group[df$education == 11] <- "Post-Grad"
df$educ_group <- factor(df$educ_group, levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"))

print(names(df))

#Removing NA/ 0 INCOME
df <- na.omit(df)
df <- df[df$total_income > 0, ]
```




# DEA - based on the smart questions in the proposal - for this survey, we can calculate state level data + individual level data, try to think in both directions - deadline - before thanksgiving (Nov 27)


1. Statistics summary, plot 

```{r }

#  Income distribution
par(mfrow = c(1, 1)) 

hist(df$total_income, 
     breaks = 100, 
     xlim = c(0, 350000),
     main = "Distribution of Annual Income",
     xlab = "Annual Income ($)",
     ylab = "Frequency",
     col = "cadetblue",
     border = "white")
```

The histogram is right-skewed, showing that most incomes are low while a few high-income outliers form a long tail to the right. Because these outliers pull the mean upward, the median is a more accurate measure of the typical income.

```{r }
# Income by Education level
boxplot(total_income ~ educ_group, data = df,
        main = "Income Ranges by Education Level",
        xlab = "Education Level",
        ylab = "Annual Income ($)",
        col = "lightblue",
        outline = FALSE, 
        las = 1) 

# Median Income
income_by_educ <- aggregate(total_income ~ educ_group, data = df, FUN = median)
print(income_by_educ)
```

The boxplots shows a clear positive link between education and income. As education levels rise, the median income consistently increases. However, the variance of salaries also gets much wider with higher education. While high school graduates earn within a narrow, consistent range, the "Post-Grad" group has a very tall box and long whiskers, indicating a huge gap between the lowest and highest earners in that category.

```{r }
# Gender income difference by education
par(mar = c(8, 4, 4, 2))
boxplot(total_income ~ gender * educ_group, data = df,
        main = "Gender Pay Gap by Education",
        xlab='',
        ylab = "Annual Income ($)",
        col = c("cornflowerblue", "pink"), # Mavi=Erkek, Pembe=Kadın
        outline = FALSE,
        las = 2, 
        cex.axis = 0.7) # Yazı boyutunu biraz küçült

legend("topleft", legend = c("Male", "Female"), 
       fill = c("cornflowerblue", "pink"), bty = "n")


```

The graph compares incomes for men and women across different education levels. It clearly shows that in every category, men have a higher median income than women. This pay gap gets much wider as education levels rise. While the difference is smaller for high school graduates, it becomes very large for the "Post-Grad" group, where men earn significantly more than women even with the same level of education.

```{r }
# Categoric
par(mfrow = c(1, 2)) 

# Education
educ_counts <- table(df$educ_group)
barplot(educ_counts, 
        main = "Sample Size by Education",
        col = "orange",
        las = 2,
        cex.names = 0.7)


```

This chart counts how many people in the study have different levels of education. The biggest group is people with a high school education or less. The smallest group is people with only some college experience.

```{r }
# Race
race_counts <- sort(table(df$race), decreasing = FALSE) 
par(mar=c(5, 6, 4, 1)) 
barplot(race_counts, 
        main = "Racial Composition", 
        xlim = c(0,100000),
        horiz = TRUE, 
        col = "lightgray",
        las = 1,
        cex.names = 0.7)
```

This chart shows the race of the people in the study. The vast majority of people are White or from the "Other Asian/Pacific" group. On the other hand, there are very few Native American or Japanese participants.


2. Normal distribution test

3. Hypothesis testing 


# ANOVA (Analysis of Variance) - FOR EDUCATION LEVELS
 
 Research Question: Does education level result in a statistically significant difference in income?
 
 H0 (Null Hypothesis): The mean income of all education groups is equal.
 
 H1 (Alternative Hypothesis): At least one group's income is different from the others.

```{r }

anova <- aov(log(total_income) ~ educ_group, data = df)

print(summary(anova))
```

A one-way ANOVA test was conducted to determine if education level creates a statistically significant difference in income. The results indicate a strong relationship, with a high F-value ($13,868$) and a p-value of less than 0.001. Because the p-value is extremely low, we reject the null hypothesis and confirm that income levels vary significantly across education groups. However, since the ANOVA does not specify which groups differ, a post-hoc test is required to identify the specific differences.

```{r }
tukey<- TukeyHSD(anova)

print(tukey)

par(mar=c(5, 15, 4, 2)) 
plot(tukey, las=1, col="red")

```

The results show that every single pairwise comparison is statistically significant. This confirms a clear hierarchy in earnings: those with 'High School or Less' earn the lowest, followed by 'Some College' and 'Bachelor's', while the 'Post-Grad' group has the highest income. In short, each step up in education level results in a statistically significant increase in total income.

# T-Test (Independent Samples) - FOR GENDER
 
 Research Question: Is there a difference between the mean incomes of males and females?
 
 H0 (Null Hypothesis): The mean incomes of males and females are EQUAL.
 
 H1 (Alternative Hypothesis): The mean incomes of males and females are NOT EQUAL.


```{r }

t_test <- t.test(log(total_income) ~ gender, data = df)
print(t_test)
```

A Welch Two Sample t-test was conducted to compare the average incomes of male and female groups. The results indicate a highly significant difference between the two genders ($t = 88, p < 0.001$). The analysis shows that the average log-income for males ($10.8$) is higher than the average log-income for females ($10.3$). Additionally, the 95% confidence interval is entirely positive ($0.46$ to $0.48$), which confirms that males consistently earn more than females in this dataset. Based on this, the null hypothesis is rejected.


# Pearson Correlation Test - income vs state gdp

Research Question: To what extent is an individual's total income associated with the economic prosperity (GDP) of their state?

H0 (Null Hypothesis): There is NO linear correlation between individual income and state-level GDP.
 
H1 (Alternative Hypothesis): There is a statistically significant linear correlation between individual income and state-level GDP.

```{r }
cor_test <- cor.test(log(df$total_income), log(df$state_real_gdp), method = "pearson")
print(cor_test)

```

A Pearson correlation test was conducted to examine the relationship between state-level GDP and individual total income. 

The results show a statistically significant relationship ($t = 14, p < 0.001$). However, the correlation coefficient is extremely low ($r = 0.03$). This indicates that although there is a positive association between the wealth of a region and personal income, the relationship is negligible. In other words, living in a richer state has almost no direct impact on an individual's income in this dataset.

# Regression

1. Plot of correlation

To explore main explanatory variable education level against main dependent variable real per capital income, we fiest look at the correlation plot between these two variables.  

Because our main explanatory variable is categorical variable, we be using correlation matrix for this section. 

For numerical variable and categorical, we need to use Eta method to determine the correlation.

```{r}
etaSquared(aov( total_income ~ education_level, data = df))
eta_value <- function(cat, num) {
  as.numeric(etaSquared(aov(num ~ cat))[1])
}

eta_value(df$education_level, df$total_income)
```

This result imply that education level explains 14.1% of the variation in total income. This is on the threshold of large effect which means education level has a meaningful and strong association with persoanl total income. 

2. Run regression

```{r linear regression without control}
model1 <- lm( total_income ~ education_level, data = df)
summary(model1)
```
Our regression estimation for simple linear regression is:

total income 2023(personal level) = 26928 + 1584*Grade5-8 + 3312*Grade9 + 3276*Grade10 + 1269*Grade11 + 12964*Grade12 + 20261*1 year College + 25400*2 year college + 57307*4 year college + 106864* 5 or more year college 


Based on the simple regression model, education have a higher and higher influence on personal level income after 12 grade and all of those coefficients are significant. This results correpsondes to the real world because law forbide child labor thus income would only be associate with education level at least after high school when a person could work legaly full time. 

```{r linear regression with control for gender, age, and race}
model2 <- lm( total_income ~ education_level + gender_code + race_code + age , data = df)
summary(model2)
```

```{r linear regression with control for gender, age, and race}
model1 <- lm( total_income ~ education_level * age + education_level * gender_code + education_level * race_code + age + race_code, data = df)
summary(model1)


```

3. Test the fit of the regression

- Confusion matrix

- Mcfaddle

- Anova Test between models (remove controlling variables)

- ROC/AUC

# Tree-based model

```{r tree 1}
set.seed(1)
income_regression <- rpart(total_income ~ education + num_children + gender + age + marital_status + race + citizenship + log(state_real_gdp) + log(state_total_jobs), data=df, method="anova", control = list(maxdepth = 7) )
```

```{r summary tree}
summary(income_regression) # detailed summary of splits
```
```{r plot tree 1}
library(rpart.plot)
rpart.plot(income_regression)
```
# Reference

Autor, D. H. (2014). Skills, education, and the rise of earnings inequality among the other 99 percent. Science, 344(6186), 843–851. https://doi.org/10.1126/science.1251868

Card, D. (1999). The causal effect of education on earnings. In O. Ashenfelter & D. Card (Eds.), Handbook of labor economics (Vol. 3, pp. 1801–1863). Elsevier.

Moretti, E. (2004). Human capital externalities in cities. In J. V. Henderson & J.-F. Thisse (Eds.), Handbook of regional and urban economics (Vol. 4, pp. 2243–2291). Elsevier. https://doi.org/10.1016/S1574-0080(04)80014-4

Ruggles, S., Flood, S., Sobek, M., Backman, D., Cooper, G., Rivera Drew, J. A., Richards, S., Rodgers, R., Schroeder, J., & Williams, K. C. W. (2025). IPUMS USA: Version 16.0 [ACS 1-year microdata for 2023]. IPUMS. https://doi.org/10.18128/D010.V16.0

U.S. Bureau of Economic Analysis. (2023). State Annual Summary Statistics: Personal Income, GDP, Consumer Spending, Price Indexes, and Employment. U.S. Department of Commerce. Retrieved from https://apps.bea.gov/regional/downloadzip.htm

