---
title: "Data Science Project 2 -  Return to Education and Income in the United States (2023)"
author:
- Constantine Zhang
- Ha Nguyen
- Kemal Ege Sural
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r library, include = FALSE}
library(tidyverse)
library(haven)
library(corrplot)
library(ggplot2)
library(reshape2)
library(lsr)
library(rpart)
library(interactions)
library(patchwork)
library(rpart.plot)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(results="markup", warning = F, message = F,
                      fig.width=8, fig.height=6, fig.dim=c(8,6), fig.align='center')
options(scientific=T, digits = 3,scipen = 999)
```

```{r data, include = FALSE}
acs2023 <- read.csv("acs2023.csv")
```


```{r }
acs_raw <- read.csv("acs2023.csv")
df <- acs_raw # copy


#Renaming

names(df)[names(df) == "inctot"]           <- "total_income"
names(df)[names(df) == "real_gdp"]         <- "state_real_gdp"
names(df)[names(df) == "real_pc_income"]   <- "state_pc_income"
names(df)[names(df) == "total_employment"] <- "state_total_jobs"

names(df)[names(df) == "educ"]     <- "education"
names(df)[names(df) == "age"]      <- "age"              
names(df)[names(df) == "sex"]      <- "gender"
names(df)[names(df) == "race"]     <- "race"
names(df)[names(df) == "marst"]    <- "marital_status"
names(df)[names(df) == "nchild"]   <- "num_children"
names(df)[names(df) == "citizen"]  <- "citizenship"      
names(df)[names(df) == "empstat"]  <- "employment_status" 
names(df)[names(df) == "labforce"] <- "labor_force"       
names(df)[names(df) == "perwt"]    <- "person_weight"

#Factor changes

df$gender <- factor(df$gender, 
                    levels = c(1, 2), 
                    labels = c("Male", "Female"))

df$marital_status <- factor(df$marital_status, levels = 1:6, 
                            labels = c("Married (Present)", "Married (Absent)", 
                                       "Separated", "Divorced", "Widowed", "Single"))


df$race <- factor(df$race, levels = 1:6,
                  labels = c("White", "Black", "Native American", 
                             "Chinese", "Japanese", "Other Asian/Pacific"))

df$citizenship <- factor(df$citizenship, levels = 1:3,
                         labels = c("Born Abroad (US Parents)", 
                                    "Naturalized Citizen", 
                                    "Not a Citizen"))

df$employment_status <- factor(df$employment_status, levels = 1:3,
                               labels = c("Employed", "Unemployed", "Not in Labor Force"))

df$labor_force <- factor(df$labor_force, levels = 1:2,
                         labels = c("Not in Labor Force", "In Labor Force"))


df$education_level <- factor(df$education, levels = 1:11,
                             labels = c("Nursery-Gr4", "Gr5-8", "Gr9", "Gr10", "Gr11", "Gr12",
                                        "1yr College", "2yr College", "3yr College", "4yr College", "5+yr College"))


df$educ_group <- NA
df$educ_group[df$education <= 6] <- "HS or Less"
df$educ_group[df$education %in% c(7, 8, 9)] <- "Some College"
df$educ_group[df$education == 10] <- "Bachelor's"
df$educ_group[df$education == 11] <- "Post-Grad"
df$educ_group <- factor(df$educ_group, levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"))

print(names(df))

#Removing NA/ 0 INCOME
df <- na.omit(df)
df <- df[df$total_income > 0, ]
```

# Literature Review & Motivation

This research project examines the economic returns to education in the United States, building on established findings in labor economics. The literature review highlights three key foundational studies:

1. Card (1999) demonstrated that higher educational attainment strongly correlates with increased earnings, with each additional year of schooling producing meaningful income gains even after accounting for demographic factors. 

2. Moretti (2004) extended this understanding by showing that the returns to education vary based on regional economic contexts, including productivity levels, industry composition, and local employment structures. 

3. Most recently, Pan (2023) provided evidence that associate degrees enable access to higher-paying firms, with approximately 6% of returns stemming from firm-specific pay premiums, suggesting that over a quarter of associate degree returns come from employment at better-compensated companies.

These studies collectively motivate an investigation into how educational returns vary across different populations and economic contexts. While the relationship between education and earnings is well-established, important questions remain about heterogeneity in returns across demographic groups and geographic regions with different economic conditions.

# Dataset Overview

## Overview:

The analysis combines two complementary data sources from 2023, yielding 229,342 observations after cleaning and merging:

1. American Community Survey (ACS) from IPUMS provides individual-level microdata including education levels, demographic characteristics, labor force status, and total income. This nationally representative dataset enables estimation of returns to education and subgroup analysis across gender, race, age, and nativity.

2. State Annual Summary Statistics from the U.S. Bureau of Economic Analysis (BEA) offers macroeconomic indicators including real GDP, GDP per capita, and total employment. These variables allow the research to control for state-level economic conditions that may influence educational returns.

## Data Limitations

The ACS data has several important constraints: 

- Cross-sectional only: cannot infer causality or follow individuals over time.
  
- Income measurement error: self-reported income, top-coding, and under-reporting among high earners.


- Coarse education categories: no detail on field of study, institution quality, degree type, or credentials.

- Small subgroup samples: certain demographic groups in smaller states have limited sample sizes, increasing variance.

The state-level economic data uses broad state averages that may obscure local labor market differences between urban and rural areas or specific industry clusters.

# Smart Questions

The project explores two primary questions:

1. ***What is the relationship between annual income and levels of education, after accounting for demographic and state-level economic covariates?*** This question seeks to quantify the income premium associated with different educational attainment levels while controlling for confounding factors.

2. ***How does the return to education vary by demographic subgroups and across states with differing economic conditions?*** This question investigates heterogeneity in educational returns, examining whether the economic benefits of education differ systematically across gender, race, age groups, and geographic regions with varying economic characteristics.

Together, these questions aim to provide a understanding of educational returns in the U.S, accounting for both individual characteristics and state economic contexts.

# EDA


## Statistics summary, plot 

```{r }

#  Income distribution
par(mfrow = c(1, 1)) 

hist(df$total_income, 
     breaks = 100, 
     xlim = c(0, 350000),
     main = "Distribution of Annual Income",
     xlab = "Annual Income ($)",
     ylab = "Frequency",
     col = "cadetblue",
     border = "white")
```

The histogram is right-skewed, showing that most incomes are low while a few high-income outliers form a long tail to the right. Because these outliers pull the mean upward, the median is a more accurate measure of the typical income.

```{r }
# Income by Education level
boxplot(total_income ~ educ_group, data = df,
        main = "Income Ranges by Education Level",
        xlab = "Education Level",
        ylab = "Annual Income ($)",
        col = "lightblue",
        outline = FALSE, 
        las = 1) 

# Median Income
income_by_educ <- aggregate(total_income ~ educ_group, data = df, FUN = median)
print(income_by_educ)
```

The boxplots shows a clear positive link between education and income. As education levels rise, the median income consistently increases. However, the variance of salaries also gets much wider with higher education. While high school graduates earn within a narrow, consistent range, the "Post-Grad" group has a very tall box and long whiskers, indicating a huge gap between the lowest and highest earners in that category.

```{r }
# Gender income difference by education
par(mar = c(8, 4, 4, 2))
boxplot(total_income ~ gender * educ_group, data = df,
        main = "Gender Pay Gap by Education",
        xlab='',
        ylab = "Annual Income ($)",
        col = c("cornflowerblue", "pink"), # Mavi=Erkek, Pembe=Kadın
        outline = FALSE,
        las = 2, 
        cex.axis = 0.7) # Yazı boyutunu biraz küçült

legend("topleft", legend = c("Male", "Female"), 
       fill = c("cornflowerblue", "pink"), bty = "n")


```

The graph compares incomes for men and women across different education levels. It clearly shows that in every category, men have a higher median income than women. This pay gap gets much wider as education levels rise. While the difference is smaller for high school graduates, it becomes very large for the "Post-Grad" group, where men earn significantly more than women even with the same level of education.

```{r }
# Categoric
par(mfrow = c(1, 2)) 

# Education
educ_counts <- table(df$educ_group)
barplot(educ_counts, 
        main = "Sample Size by Education",
        col = "orange",
        las = 2,
        cex.names = 0.7)


```

This chart counts how many people in the study have different levels of education. The biggest group is people with a high school education or less. The smallest group is people with only some college experience.

```{r }
# Race
race_counts <- sort(table(df$race), decreasing = FALSE) 
par(mar=c(5, 6, 4, 1)) 
barplot(race_counts, 
        main = "Racial Composition", 
        xlim = c(0,100000),
        horiz = TRUE, 
        col = "lightgray",
        las = 1,
        cex.names = 0.7)
```

This chart shows the race of the people in the study. The vast majority of people are White or from the "Other Asian/Pacific" group. On the other hand, there are very few Native American or Japanese participants.

## Hypothesis Testing

### ANOVA (Analysis of Variance) - FOR EDUCATION LEVELS
 
 Research Question: Does education level result in a statistically significant difference in income?
 
 H0 (Null Hypothesis): The mean income of all education groups is equal.
 
 H1 (Alternative Hypothesis): At least one group's income is different from the others.

```{r }

anova <- aov(log(total_income) ~ educ_group, data = df)

print(summary(anova))
```

A one-way ANOVA test was conducted to determine if education level creates a statistically significant difference in income. The results indicate a strong relationship, with a high F-value ($13,868$) and a p-value of less than 0.001. Because the p-value is extremely low, we reject the null hypothesis and confirm that income levels vary significantly across education groups. However, since the ANOVA does not specify which groups differ, a post-hoc test is required to identify the specific differences.

```{r }
tukey<- TukeyHSD(anova)

print(tukey)

par(mar=c(5, 15, 4, 2)) 
plot(tukey, las=1, col="red")

```

The results show that every single pairwise comparison is statistically significant. This confirms a clear hierarchy in earnings: those with 'High School or Less' earn the lowest, followed by 'Some College' and 'Bachelor's', while the 'Post-Grad' group has the highest income. In short, each step up in education level results in a statistically significant increase in total income.

### T-Test (Independent Samples) - FOR GENDER
 
 Research Question: Is there a difference between the mean incomes of males and females?
 
 H0 (Null Hypothesis): The mean incomes of males and females are EQUAL.
 
 H1 (Alternative Hypothesis): The mean incomes of males and females are NOT EQUAL.


```{r }

t_test <- t.test(log(total_income) ~ gender, data = df)
print(t_test)
```

A Welch Two Sample t-test was conducted to compare the average incomes of male and female groups. The results indicate a highly significant difference between the two genders ($t = 88, p < 0.001$). The analysis shows that the average log-income for males ($10.8$) is higher than the average log-income for females ($10.3$). Additionally, the 95% confidence interval is entirely positive ($0.46$ to $0.48$), which confirms that males consistently earn more than females in this dataset. Based on this, the null hypothesis is rejected.


### Pearson Correlation Test - income vs state gdp

Research Question: To what extent is an individual's total income associated with the economic prosperity (GDP) of their state?

H0 (Null Hypothesis): There is NO linear correlation between individual income and state-level GDP.
 
H1 (Alternative Hypothesis): There is a statistically significant linear correlation between individual income and state-level GDP.

```{r }
cor_test <- cor.test(log(df$total_income), log(df$state_real_gdp), method = "pearson")
print(cor_test)

```

A Pearson correlation test was conducted to examine the relationship between state-level GDP and individual total income. 

The results show a statistically significant relationship ($t = 14, p < 0.001$). However, the correlation coefficient is extremely low ($r = 0.03$). This indicates that although there is a positive association between the wealth of a region and personal income, the relationship is negligible. In other words, living in a richer state has almost no direct impact on an individual's income in this dataset.

# Regression

##  Plot of correlation

To explore the main explanatory variable, education level, against the main dependent variable, real per capita income, we first look at the correlation plot between these two variables.

Because our main explanatory variable is categorical, we do not use a standard correlation matrix for this section.

For a numerical variable and a categorical variable, we use the eta method to determine the correlation.

```{r}
etaSquared(aov( total_income ~ education_level, data = df))
eta_value <- function(cat, num) {
  as.numeric(etaSquared(aov(num ~ cat))[1])
}

eta_value(df$education_level, df$total_income)
```

This result implies that education level explains 14.1% of the variation in total income. This is on the threshold of a large effect, which means education level has a meaningful and strong association with personal total income.

##  Run regression

```{r change level to factor and reference level}
df$education_level <- factor(df$educ_group)
df$race <- factor(df$race)
df$gender <- factor(df$gender)
df$race <- relevel(df$race, ref = "White")
df$gender <- relevel(df$gender, ref = "Male")
df$educ_group <- relevel(df$educ_group, ref = "HS or Less")
```

### linear regression without control

```{r linear regression without control}
model.base <- lm( log(total_income) ~ educ_group, data = df)
summary(model.base)
```

### linear regression with control for gender, age, and race

```{r linear regression with control for gender, age, and race}
model.gender.race <- lm( log(total_income) ~ educ_group + gender + race + age , data = df)
summary(model.gender.race)
```
This specification, with age, gender, and race as controls, has a higher R-squared value. We also find that all coefficients are statistically significant. However, this specification only allows differences in intercepts, not slopes, across race and gender.

Based on the simple regression model, education has an increasingly strong influence on personal income with each additional level of education a person acquire, and all of these coefficients are statistically significant. 


### linear regression with interaction terms for gender, age, and race


```{r linear regression with interaction terms for gender, age, and race}
model.interaction <- lm( log(total_income) ~  educ_group   * gender + educ_group* race + gender * race +  +age  , data = df)
summary(model.interaction)
```

Our third regression model specification allows the slopes to vary across race and gender. The R-squared of this model is slightly higher, and we obtain statistical significance for some coefficients due to the substantially large number of observations in the data frame.

### linear regression with interaction terms for gender, age, and race, and macro control


```{r state level macro}
model.interaction.state <- lm( log(total_income) ~  educ_group   * gender + educ_group* race + gender * race + age + log(state_total_jobs) + log(state_pc_income) + log(state_real_gdp), data = df)
summary(model.interaction.state)
```
### VIF selection 

```{r vif check model with control}
car::vif(model.gender.race)
```

```{r vif check model with interactions}
car::vif(model.interaction)
```

```{r vif check model with interactions and state control}
car::vif(model.interaction.state)
```
From the VIF checks of the three models, we found that the state-level control variables—total number of jobs and real GDP—exhibit multicollinearity. We therefore decide to keep total jobs and remove real GDP from our final specification.

## final model

```{r final model}
model.interaction.state <- lm( log(total_income) ~  educ_group   * gender + educ_group* race + gender * race + age + log(state_total_jobs) + log(state_pc_income) , data = df)
summary(model.interaction.state)
```
Final model consists of interaction terms for education level, gender, and race. The purpose of these interaction terms is to allow the regression model to have different slopes, thereby changing the fitted values across racial and gender groups. 

Age is included to account for the fact that income is usually positively correlated with age. The state-level macroeconomic control variables account for differences in economic activity across states.

The regression results show that education has a strong positive effect on income. Females earn less than males on average, although the gender gap varies across racial groups. Gaps are smaller for Black, Chinese, and Other Asian/Pacific individuals, but slightly larger for Japanese. Race also significantly affects income: predicted earnings are highest for White and Japanese individuals and lowest for Black and Native American individuals. Interactions between education and race indicate that returns to education are strongest for Chinese and Other Asian/Pacific groups, while slightly lower for Black and Native American groups. Age and state-level economic conditions, including total jobs and per capita income, are positively associated with income.

Overall, income rises with education level, but returns of education are different among each racial group. For example: we estimate that a white male person who finish a bachelor's degree would see their income rise by 79%. However, for a black male person, we estimate only about 50.2% increase in their income. 

```{r}
df$educ_group <- factor(df$educ_group,
                        levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"),
                        ordered = TRUE)
levels(df$educ_group)


```

## Interaction  Plots
To better understanding and interpreting the result coefficients, we employ interaction plots for the third specification. 

```{r interaction plot gender}
newdata <- expand.grid(
  educ_group = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"),
  gender = unique(df$gender),
  race = unique(df$race),
  age = mean(df$age, na.rm = TRUE),
  state_total_jobs = mean(df$state_total_jobs, na.rm = TRUE),
  state_pc_income = mean(df$state_pc_income, na.rm = TRUE),
  state_real_gdp = mean(df$state_real_gdp, na.rm = TRUE)
)


newdata$educ_group <- factor(newdata$educ_group,levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"), ordered = TRUE)
predictions <- predict(model.interaction.state, newdata, se.fit = TRUE, interval = "confidence")
pred_df <- cbind(newdata, predictions)

#colnames(pred_df)
pred_df <- pred_df %>%
  rename(
    log_income_pred = fit.fit,
    ci_lower = fit.lwr,
    ci_upper = fit.upr
  )
pred_summary <- pred_df %>%
  group_by(educ_group, gender) %>%
  summarise(
    log_income_pred = mean(log_income_pred),
    ci_lower = mean(ci_lower),
    ci_upper = mean(ci_upper),
    .groups = "drop"
  )


ggplot(pred_summary, aes(x = educ_group, y = log_income_pred, color = gender)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
                width = 0.2,
                position = position_dodge(width = 0.5)) +
  scale_x_discrete(limits = c("HS or Less", "Some College", "Bachelor's", "Post-Grad")) +
  labs(x = "Education Level",
       y = "Predicted log(Income)",
       color = "Gender",
       title = "Predicted log(Income) by Education and Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This interaction plot represents the predicted values of log income for males and females based on their education levels and includes 95% confidence intervals. This graph shows the expected results: as education level increases, income increases. Moreover, males have higher income than females at all education levels.

```{r interaction plot race }
library(emmeans)
race_emm <- emmeans(model.interaction.state, ~ race)  
race_df <- as.data.frame(race_emm)
#colnames(race_df)
ggplot(race_df, aes(x = race, y = emmean)) +
  geom_point(size = 3, color = "darkgreen") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, color = "darkgreen") +
  labs(x = "Race",
       y = "Predicted log(Income)",
       title = "Predicted log(Income) by Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

This interaction plot demonstrates the predicted values of log income by race. The results are also expected, as White individuals have the highest predicted income, while Native American and African American individuals have the lowest predicted income.

```{r race and gender plot}
race_gender_emm <- emmeans(model.interaction.state, ~ race * gender)
race_gender_df <- as.data.frame(race_gender_emm)

race_gender_df$race <- factor(race_gender_df$race, levels = levels(df$race))

ggplot(race_gender_df, aes(x = race, y = emmean, color = gender)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.2,
                position = position_dodge(width = 0.5)) +
  labs(x = "Race",
       y = "Predicted log(Income)",
       color = "Gender",
       title = "Predicted log(Income) by Race and Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Next, we include a graph of predicted log income based on both race and gender to examine the gender wage gap across different races. Japanese individuals have the largest gender income gap, followed by White individuals, while Black and Chinese individuals have the smallest gender income gap.

```{r race, gender, and education level}
pred_emm <- emmeans(model.interaction.state, ~ educ_group * gender * race)
pred_df <- as.data.frame(pred_emm)


pred_df$educ_group <- factor(pred_df$educ_group,
                             levels = c("HS or Less", "Some College", "Bachelor's", "Post-Grad"),
                             ordered = TRUE)
pred_df$race <- factor(pred_df$race, levels = levels(df$race))
pred_df$gender <- factor(pred_df$gender, levels = levels(df$gender))

ggplot(pred_df, aes(x = educ_group, y = emmean, color = gender)) +
  geom_point(position = position_dodge(width = 0.5), size = 1.5) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.2,
                position = position_dodge(width = 0.5)) +
  facet_wrap(~ race) +
  labs(x = "Education Level",
       y = "Predicted log(Income)",
       color = "Gender",
       title = "Predicted log(Income) by Education, Gender, and Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
This interaction plot combines all factors, showing predicted log income based on gender, race, and education level. It is clear that log income and education have a positive correlation within every racial group, and the gender income gap persists across all racial groups. 


3. Test the fit of the regression

- Anova Test between models (remove controlling variables)

```{r test for base model and control variable model}
anova(model.base, model.gender.race)
```
An ANOVA test comparing the base model to the model with control variables shows a p-value close to 0, indicating that the control variables—age, gender, and race—significantly improve the model.


```{r test control model and interaction model}
anova(model.gender.race, model.interaction)
```
An ANOVA test comparing the model with only control variables to the model with interaction terms shows a p-value close to 0, indicating that adding interaction terms and allowing slopes to vary across race and gender significantly improves the model.

```{r test state control model and interaction model }
anova(model.interaction, model.interaction.state)
```

# Tree-based model

## Regression tree (restricted version)

### Grow the tree

**Model Specification**: 

To investigate the relationship between educational attainment and income, we employed a regression tree model using the Classification and Regression Trees (CART) algorithm. The model predicts total annual income as a continuous outcome through recursive binary partitioning that minimizes mean squared error at each split.

**Variables and Controls**: 

- Predictors included:

  - Demographics: educational level (categorical, as we want to examine each level of education attainment will impact the total income), gender, age, and race
  
  - State-level macroeconomic controls: log-transformed real GDP, per capita income, and total employment to account for regional economic variations

- Model Parameters

  - Maximum depth: 7 levels to balance complexity with interpretability
  
  - Complexity parameter (CP): 0.01 (default) for cost-complexity pruning, requiring each split to reduce prediction error by at least 1%

  
```{r tree 1}
set.seed(1)

df <- df %>%
  mutate(education_cat = factor(education))

income_restricted <- rpart(total_income ~ education_cat + gender + age + race + log(state_real_gdp) + log(state_pc_income) + log(state_total_jobs), data=df, method="anova", control = list(maxdepth = 7) )
```

```{r summary tree}
summary(income_restricted) 
```
**Model Summary interpretation**:

1. Complexity Parameter (CP) Table

Cross-validation results show strong performance:

- The model made 3 splits before reaching the CP = 0.01 threshold

- Relative error decreased from 1.000 to 0.844, representing a 15.6% reduction in prediction error

- Cross-validation error (xerror) closely tracks training error, indicating the model generalizes well without overfitting

- Each successive split provides diminishing returns (10.8% → 3.0% → 1.7% → 1.0% reduction)

The tree stopped at 3 splits because the 4th split barely met the 1% improvement threshold, and no subsequent splits would have provided sufficient error reduction to justify additional complexity.

2. Variable Importance
Dramatic dominance of education:

- Education: 89% of total variable importance

- Gender: 11% of total variable importance

- All other variables (age, race, state controls): 0% - did not contribute to any splits

```{r plot tree 1}
rpart.plot(
  income_restricted,
  type = 2,          
  extra = 101,       
  under = TRUE,      
  faclen = 0,        
  digits = -3,
  tweak = 1.1        
)
```
**Plot interpretation**

1. Primary Split

- Root node split: Education is the single most important predictor

- Left branch (75.2% of sample, n=172,509): Education levels 1-10 ($\leq$ 4 years of college). Average predicted income: $56,676

- Right branch (24.8% of sample, n=56,833): Education levels 11 (higher educational attainment, 5+ years of college). Average predicted income: $133,792

There's a dramatic income divide (~$77,000 difference) between those with education levels 1-10 versus 11, representing the clearest earnings premium in the data.

2. Secondary Splits

- Left Branch (Lower Education): Further Split by Education: The lower education group subdivides again by education:

  - Education levels 1-8, meaning $\leq$  2 years of college (48.9% of total sample): $41,809 average income
  
  - Higher levels within this group (10 - 4 years of college) (26.4% of total sample): $84,235 average income

This indicates meaningful income gradations even within the lower education categories.

3. Right Branch (Higher Education): Split by Gender

The higher education group subdivides by gender:

- Females (11.4% of total sample): $104,968 average income

- Males (13.4% of total sample): $158,407 average income

Among the highest educated individuals, a substantial gender pay gap exists (~$53,000 difference), suggesting gender disparities are most pronounced at higher education levels.

**Overall Takeaway**

1. Education is the overwhelmingly strongest predictor of income: It defines the first split and most of the structure.

2. Gender is the second-most important variabl: Especially for high education groups.

3. State-level economics barely affect individual income.

### Prune the Tree

We can "prune" the tree to reduce variance. A plot can be made after pruning, showing the effects in terms of deviance or other statistics.

```{r prune}
library(rattle)

print(income_restricted$cptable)
plotcp(income_restricted)
income_prune<- prune(income_restricted, cp = income_restricted$cptable[3,"CP"])
rpart.plot(
  income_prune,
  type = 2,          
  extra = 101,       
  under = TRUE,      
  faclen = 0,        
  digits = -3,
  tweak = 1.1        
)
```
The pruned tree is simpler than the original:

- Depth reduced: From 3 levels to 2 levels (one fewer split)

- Terminal nodes: 3 nodes instead of 4

- Gender split removed: The gender distinction within high-education individuals has been pruned away

The tree reveals a simple three-level income hierarchy based purely on education:

- Lowest tier (48.9%): ~$42K - Education levels 1-8

- Middle tier (26.4%): ~$84K - Higher levels education 10

- Highest tier (24.8%): ~$134K - Education levels 11

### Train/Test Split with Tree Pruning Evaluation”

We split the dataset into 5 random groups. Pick one group as test set in turns, and the remaining four as training data, we can build five different models and score the models using the test data five times. 

1. First fold Cross-Validation (1-fold CV) for Pruning

```{r, results="markup"}
set.seed(1)
df$fold <- floor(runif(nrow(df), 1, 5))
test.set <- df[df$fold == 1, ]
train.set <- df[df$fold != 1, ]
```

```{r train}
treefit_fold <- rpart(total_income ~ education_cat + gender + age + race +
                        log(state_real_gdp) + log(state_pc_income) + log(state_total_jobs),
                      data = train.set,
                      method = "anova",
                      control = list(maxdepth = 7))
rpart.plot(
  treefit_fold,
  type = 2,          
  extra = 101,       
  under = TRUE,      
  faclen = 0,        
  digits = -3,
  tweak = 1.1        
)
```

This train set regression tree predicts total income using demographic and economic variables from 171,868 training observations. The tree maintains identical structure: education splits first (categories 1-10), followed by gender within the highest education group. Predicted incomes remain nearly identical to the full tree.

```{r predict}
pred <- predict(treefit_fold, test.set)
plot(pred,test.set$total_income, main="Predicted vs. Observed Total Income - Test set", xlab="Tree Predicted Salary", ylab="Observed Salary")
treefit_fold_error <- sum((pred - test.set$total_income)^2)
```

```{r prune the tree}
treefit_fold_prune <- prune(treefit_fold,
                            cp = treefit_fold$cptable[3, "CP"])

pred2 <- predict(treefit_fold_prune, test.set)
plot(pred2,test.set$total_income, main="Predicted vs. Observed Total Income - Test set - Pruned", xlab="Tree Predicted Salary", ylab="Observed Salary")
treefit_fold_prune_error <- sum((pred2 - test.set$total_income)^2)
```

This scatter plot reveals the model's fundamental limitation at both tree. For the full tree (restricted version),  it creates four distinct income predictions, actual incomes within each group vary enormously. For the pruned tree,  it creates only 4 distinct income predictions. However, individuals assigned the same prediction earn anywhere from \$0 to \$500,000+. The extreme vertical dispersion demonstrates that education and gender alone cannot predict individual income with useful precision. 

2. Testing all 5 folds

```{r four fold cv}
library(rpart.plot)

set.seed(1)

K <- 5
df$fold_2 <- sample(1:K, size = nrow(df), replace = TRUE)

cv_results <- data.frame(
  fold         = 1:K,
  full_error   = NA_real_,
  pruned_error = NA_real_,
  cp_row3      = NA_real_,
  error_diff   = NA_real_,   
  compare      = NA_character_   
)

for (k in 1:K) {
  test.set  <- df[df$fold_2 == k, ]
  train.set <- df[df$fold_2 != k, ]
  
  treefit_k <- rpart(
    total_income ~ education_cat + gender + age + race + 
                        log(state_real_gdp) + log(state_pc_income) + log(state_total_jobs),
    data   = train.set,
    method = "anova",
    control = rpart.control(maxdepth = 7)
  )
  
  pred_full <- predict(treefit_k, newdata = test.set)
  sse_full  <- sum((pred_full - test.set$total_income)^2)
  cv_results$full_error[k] <- sse_full
  
  cp_row3 <- treefit_k$cptable[3,"CP"]
  cv_results$cp_row3[k] <- cp_row3
  
  treefit_k_pruned <- prune(treefit_k, cp = cp_row3)
  
  pred_pruned <- predict(treefit_k_pruned, newdata = test.set)
  sse_pruned  <- sum((pred_pruned - test.set$total_income)^2)
  cv_results$pruned_error[k] <- sse_pruned

  cv_results$error_diff[k] <- sse_pruned - sse_full
  
  if (sse_pruned < sse_full) {
    cv_results$compare[k] <- "Pruned < Full"   
  } else if (sse_pruned > sse_full) {
    cv_results$compare[k] <- "Pruned > Full"   
  } else {
    cv_results$compare[k] <- "Pruned = Full"  
  }
}


cv_results
```
Across all five folds of the cross-validation exercise, the pruned tree consistently produced a higher sum of squared errors than the full tree, indicating that pruning reduced predictive accuracy in this dataset. This outcome is expected because the fully grown tree is already extremely small, with only three splits.

We selected the subtree associated with the smallest xstd in the complexity table, leading to choosing the third row. Although this row provides the most stable cross-validated error estimate, it also corresponds to a simpler tree that eliminates the final split present in the full model. 

By removing an informative split, the pruned tree loses explanatory power and underfits relative to the full model. As a result, the pruned trees capture less variation in total income, leading to systematically worse out-of-sample performance. In this setting, pruning does not improve generalization; the full tree provides the better balance of fit and complexity for predictive accuracy.

Summary table across folds:

```{r error_summary}
error_summary <- data.frame(
  model    = c("Full tree", "Pruned tree"),
  mean_SSE = c(mean(cv_results$full_error),
               mean(cv_results$pruned_error))
)

error_summary
```

The summary table of average SSE across folds reinforces the cross-validation results: the full tree achieves a lower mean SSE compared to the pruned tree, indicating that the pruned model performs worse on average. This outcome aligns with the individual-fold results, where pruning consistently increased prediction error. 

Because the full tree in this dataset is already extremely small, further pruning-based on selecting the subtree with the smallest xstd-removes an informative split and forces the model into an overly simplified structure. As a result, the pruned tree systematically underfits and fails to explain as much variation in total income as the full tree. Overall, the mean SSE comparison confirms that pruning does not improve predictive accuracy in this case, and the full tree provides better out-of-sample performance.

## Tree (unrestricted version)

### Grow the tree

At the last tree, we set CP to be 0.01, which is the default and restricted value. We now relaxed all the condition to expand the tree

```{r less restricted tree}
income_full <- rpart(
  total_income ~ education_cat + gender + age + race +
    log(state_real_gdp) + log(state_pc_income) + log(state_total_jobs),
  data   = df,
  method = "anova",
  control = rpart.control(
    maxdepth = 7,  
    cp       = 0.001,
    minsplit = 6000,   
    minbucket = 2000    
  )
)
```

```{r less restrict tree png, include = FALSE}
png("income_tree_unrestricted.png", width = 3000, height = 2400, res = 300)

rpart.plot(
  income_full,
  type = 2,
  extra = 101,
  under = TRUE,
  faclen = 0,
  digits = -3,
  tweak = 1.1
)

dev.off()
```
```{r less restrict tree}
rpart.plot(
  income_full,
  type = 2,
  extra = 101,
  under = TRUE,
  faclen = 0,
  digits = -3,
  tweak = 1.1
)
```

Takeaways from the unrestricted tree:

- Education remains the strongest predictor (first split).

- Now age, gender, and race show up repeatedly across branches.

- State macro variables appear now, especially among high-education/high-income groups.

- Final leaves show richer segmentation of the population.

This means that the unrestricted tree can detect heterogeneity in income patterns that the restricted model is too shallow to capture.

### Evaluation of restricted and unrestricted tree

```{r observed actual calculation}

set.seed(1)
df$fold_3 <- sample(1:5, size = nrow(df), replace = TRUE)
test.set_unr  <- df[df$fold_3 == 1, ]
train.set_unr <- df[df$fold_3 != 1, ]


income_full_fold <- rpart(
  total_income ~ education_cat + gender + age + race +
    log(state_real_gdp) + log(state_pc_income) + log(state_total_jobs),
  data   = train.set_unr,
  method = "anova",
  control = rpart.control(
    maxdepth  = 7,
    cp        = 0.001,
    minsplit  = 6000,
    minbucket = 2000
  )
)
```

```{r plotting}
pred_unr <- predict(income_full_fold, newdata = test.set_unr)

leaf_means  <- sort(unique(pred_unr))          
leaf_id_num <- match(pred_unr, leaf_means)     
leaf_factor <- factor(
  leaf_id_num,
  labels = round(leaf_means, 0)
)

x_numeric <- as.numeric(leaf_factor)

plot(
  jitter(x_numeric, amount = 0.15),
  test.set_unr$total_income,
  xlab = "Terminal Node (Leaf ID)",
  ylab = "Observed Total Income",
  main = "Predicted vs. Observed Total Income - Test set - Unrestricted tree",
  pch = 16,
  col = rgb(0,0,1,0.4),
  xaxt = "n"
)

axis(1, at = unique(x_numeric), labels = levels(leaf_factor))
sse_unr_test <- sum((pred_unr - test.set_unr$total_income)^2)
```
```{r compute sse}

SSE_restricted_full <- sum((df$total_income - predict(income_restricted))^2)


SSE_restricted_pruned <- sum((df$total_income - predict(income_prune))^2)

SSE_unrestricted_full <- sum((df$total_income - predict(income_full))^2)

model_compare_full <- data.frame(
  Model = c("Restricted Tree", "Restricted Pruned Tree", "Unrestricted Tree"),
  SSE   = c(SSE_restricted_full, SSE_restricted_pruned, SSE_unrestricted_full)
)

model_compare_full
```

The unrestricted tree demonstrates that adding complexity doesn't solve the prediction problem—it merely slices the unexplained variance into more granular buckets. The SSE improvement is achieved by creating 25 specific prediction values instead of 4, but individuals within each new prediction group still exhibit enormous income variation (showed in the scatter plot above). This proves the issue is missing variables (occupation, hours worked, employment status, experience), not insufficient model complexity.

# Conclusion

This analysis of 229,342 individuals from the 2023 American Community Survey demonstrates that education is the strongest predictor of income in the United States, with each successive educational level—from high school to some college, bachelor's degree, and post-graduate education—yielding statistically significant income gains confirmed across ANOVA, regression models, and regression tree analysis. 

Three critical findings emerge:

- First, large pay gaps remain: gender and race disparities persist at all education levels, with men earning significantly more than women, and a \$52,000 gender pay gap appearing among the most educated workers. 

- Second, state economic conditions play almost no role: individual income shows only weak correlation with state GDP, employment, and other macroeconomic indicators, suggesting that granular county-level data capturing local labor markets, cost of living, and industry clusters may have stronger predictive power than broad state averages. 

- Third, all models agree but explain limited variance: linear regression, ANOVA, and tree-based methods consistently highlight education first and gender/race second, yet achieve only 15-18% explanatory power ($R^2$). Comparison between simple (4 nodes) and complex (25+ nodes) tree models reveals this is a data limitation rather than a modeling problem: both show massive within-group variance where individuals assigned to the same predicted income group have actual earnings ranging from \$0 to \$500,000+, indicating that unmeasured factors such as occupation, hours worked, employment status, and experience may drive the majority of income variation


# Reference

Card, D. (1999). The causal effect of education on earnings. In O. Ashenfelter & D. Card (Eds.), Handbook of labor economics (Vol. 3, pp. 1801–1863). Elsevier. https://doi.org/10.1016/S1573-4463(99)03011-4

Moretti, E. (2004). Human capital externalities in cities. In J. V. Henderson & J.-F. Thisse (Eds.), Handbook of regional and urban economics (Vol. 4, pp. 2243–2291). Elsevier. https://doi.org/10.1016/S1574-0080(04)80014-4

Pan, W. (2025). Firm-specific pay premia and the returns to higher education: Evidence from community colleges. Labour Economics, 93, Article 102685. https://doi.org/10.1016/j.labeco.2025.102685

Ruggles, S., Flood, S., Sobek, M., Backman, D., Cooper, G., Rivera Drew, J. A., Richards, S., Rodgers, R., Schroeder, J., & Williams, K. C. W. (2025). IPUMS USA: Version 16.0 [ACS 1-year microdata for 2023]. IPUMS. https://doi.org/10.18128/D010.V16.0

U.S. Bureau of Economic Analysis. (2023). State Annual Summary Statistics: Personal Income, GDP, Consumer Spending, Price Indexes, and Employment. U.S. Department of Commerce. Retrieved from https://apps.bea.gov/regional/downloadzip.htm


